Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies

### 2025-12-31 manuscript/manuscript.qmd
Add epigraph + some other notes

---
title: "Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies"
short-title: "Clarifying Correlation and Causation"
date: 2024-11-22
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  url: https://aysps.gsu.edu/phd-student/ye-meng/
  orcid: 0000-0002-6552-8338
correspondence-prefix: "Correspondence concerning this article should be addressed to"
published: Drafting
abstract: |
  Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science. Randomized experiments have long served as a gold standard for making valid causal inferences, but most of the data social scientists work with is observational and non-experimental. However, with newer methodological developments in economics, political science, epidemiology, and other disciplines, an increasing number of studies in social science make causal claims with observational data. As a newer interdisciplinary field, however, nonprofit studies has lagged behind other disciplines in its use of observational causal inference. In this paper, we present a hands-on introduction and guide to design-based observational causal inference methods. We first review and categorize all studies making causal claims in top nonprofit studies journals over the past decade to illustrate the field’s current of experimental and observational approaches to causal inference. We then introduce a framework for modeling and identifying causal processes using directed acyclic graphs (DAGs) and provide a walk-through of the assumptions and procedures for making inferences with a range of different methods, including matching, inverse probability weighting, difference-in-differences, regression discontinuity designs, and instrumental variables. We illustrate each approach with synthetic and empirical examples and provide sample R and Stata code for implementing these methods. We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
thanks: |
  Acknowledgments here if any
epigraph:
  - text: "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them."
    source: "Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2"
keywords:
  - those
  - go
  - here
word-count: ""
bibliography: references.json
link-citations: true

@@

@@

@@

@@

>>>>

---
title: "Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies"
short-title: "Clarifying Correlation and Causation"
date: 2025-04-24
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  url: https://aysps.gsu.edu/phd-student/ye-meng/
  orcid: 0000-0002-6552-8338
correspondence-prefix: "Correspondence concerning this article should be addressed to"
published: Drafting
abstract11: |
  Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science. Randomized experiments have long served as a gold standard for making valid causal inferences, but most of the data social scientists work with is observational and non-experimental. However, with newer methodological developments in economics, political science, epidemiology, and other disciplines, an increasing number of studies in social science make causal claims with observational data. As a newer interdisciplinary field, however, nonprofit studies has lagged behind other disciplines in its use of observational causal inference. In this paper, we present a hands-on introduction and guide to design-based observational causal inference methods. We first review and categorize all studies making causal claims in top nonprofit studies journals over the past decade to illustrate the field’s current of experimental and observational approaches to causal inference. We then introduce a framework for modeling and identifying causal processes using directed acyclic graphs (DAGs) and provide a walk-through of the assumptions and procedures for making inferences with a range of different methods, including matching, inverse probability weighting, difference-in-differences, regression discontinuity designs, and instrumental variables. We illustrate each approach with synthetic and empirical examples and provide sample R and Stata code for implementing these methods. We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
thanks: |
  Acknowledgments here if any
keywords11:
  - those
  - go
  - here
word-count: ""
bibliography: references.json
link-citations: true

@@

{{< epigraph "Scientia et potentia humana in idem coincidunt, quia ignoratio causae destituit effectum. *(Human knowledge and human power meet in one; for where the cause is not known the effect cannot be produced.)*" source="Francis Bacon, *Novum Organum*, part I, aphorism III" >}}

{{< epigraph "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them." source="Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2" >}}

@@

Qualitative case studies: Rohlfing:2012, GeorgeBennett:2005, BennettCheckel:2015

@@

From thread by Stephen Wild (https://bsky.app/profile/stephenjwild.bsky.social/post/3ldhouhkcsk2z)

- Causal inference is hard: https://www.nature.com/articles/s41562-024-01939-z
- Always clearly define your estimand: https://journals.sagepub.com/doi/10.1177/00031224211004187
- Threats to validity (with DAGs): https://journals.lww.com/epidem/fulltext/2020/05000/A_Graphical_Catalog_of_Threats_to_Validity_.11.aspx
- Crash course in good and bad controls: https://journals.sagepub.com/doi/full/10.1177/00491241221099552
- Need to think about causality when selecting control variables: https://journals.sagepub.com/doi/full/10.1177/25152459221095823
- Conditioning on colliders is generally bad: https://www.annualreviews.org/content/journals/10.1146/annurev-soc-071913-043455

@@

For example, in 2022, using a large national-level observational dataset, researchers found an association between smoking behavior and reduced COVID-19 risk: people who smoked regularly saw a sigificantly reduction in the probability of contracting severe COVID-19. However, this paradoxical finding can be explained with collider bias [@AuYeungKwok:2022]. The dataset contained information about hospital patients who were (1) smokers diagnosed with COVID, (2) smokers diagnosed with smoking-related respiratory disease, and (3) non-smokers diagnosed with COVID. The data did not contain any patients who were non-smokers diagnosed with non-COVID, smoking-related respiratory disease, as these people were not patients and needed no medical treatment. In DAG terms, presence at the hospital was a collider (Smoking → Hospital ← Severe COVID), and conditioning the estimate of Smoking → Severe COVID on being a patient therefore distorted and biased the measured effect.
